\school{University of Oregon, Eugene, OR, USA}
\school{California Polytechnic State University, San Luis Obispo, CA, USA}

\degree{Doctor of Philosophy, Physics, 2022, University of Oregon}
\degree{Bachelor of Science, Physics, 2016, Cal Poly San Luis Obispo}

\interests{Gravitational Wave Astrophysics}

\position{Research Assistant, Department of Physics, University of Oregon, 2017-present}
\position{Teaching Assistant, Department of Physics, University of Oregon, 2016-2019}
\position{} % extra empty position is unnecessary

% \award{LIGO Scientific Collaboration Fellow, LIGO Hanford Observatory, 2017}
% \award{} % extra empty award is necessary

\publication{Abbott, R., et al. (2021). Search for intermediate mass black hole binaries in the third observing run of Advanced LIGO and Advanced Virgo. \emph{Astronomy \& Astrophysics}, 659, A84.}
\publication{Abbott, R., et al. (2022). Search for Gravitational Waves Associated with Gamma-Ray Bursts Detected by Fermi and Swift during the LIGO–Virgo Run O3b. \emph{The Astrophysical Journal}, 928(2), 186.}
\publication{Buikema, A., et al. (2020). Sensitivity and performance of the Advanced LIGO detectors in the third observing run. \emph{Physics Review D}, 102, 062003.}
\publication{Davis, D., et al. (2021). LIGO detector characterization in the second and third observing runs. \emph{Classical Quantum Gravity}, 38(13), 135014.}
\publication{Fiori I., Effler A., Nguyen P., Paoletti F., Schofield R.M.S., Tringali M.C. (2021) Environmental Noise in Gravitational-Wave Interferometers. In: Bambi C., Katsanevas S., Kokkotas K.D. (eds) Handbook of Gravitational Wave Astronomy. Springer, Singapore.}
\publication{Nguyen, P., et al. (2021). Environmental noise in Advanced LIGO detectors. \emph{Classical Quantum Gravity}, 38(14), 145001.}
\publication{} % extra empty publication is necessary
